Fourier-domain modulations and delays of gravitational-wave signals

### 2018-06-18 docs/paper_method.tex
Edits to the paper.

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation, extending them by including delays and higher-order corrections while being applicable to a broader class of signals including full inspiral-merger-ringdown waveforms. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving rise to limited unfaithfulness. We further present an alternative direct convolution approach to accurately represent these post-merger features.

@@

\jgb{With the unprecedented recent gravitational-wave detections of coalescencing binary black holes and binary neutron stars, announced by the LIGO-Virgo collaboration~\cite{LIGO-theevent-2016,LIGO-christmasevent-2016,LIGO-O1BBH-2016}[update refs], gravitational-wave astronomy has entered its observational era. As LIGO prepares for even more sensitive observation runs, and with the recent expansion of the ground-based detectors network with Virgo~\cite{Virgo} ( and eventually also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), observations of such compact object coalescences are expected at ever-increasing rate, with binary black holes likely to be the dominant sources~\cite{} \SM{[C]}.}

  \jgb{Moreover, the European Space Agency has recently selected the Laser Interferometer Space Antenna (LISA)\cite{LISA17} to realize the ``Gravitational Universe'' science theme\cite{GravitationalUniverse2013} as the 3rd large space mission of its Cosmic Vision program, with a tentative launch around 2034. LISA will be able to detect and characterize, among several important gravitational wave source targets, comparable-mass binary black hole coalescences from cosmological distances over a wide range of masses. These will range from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.}

  Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution. \SM{[objective: accuracy, not necessary necessary for LIGO now but in the future yes]} The greater sensitivity of LISA and other future instruments will require further increases in the accuracy and computational efficiency of signal templates.

  %This motivates an important and ongoing effort to improve gravitational wave templates. Two important focus areas for template improvements are need to cover all phases of the evolution of the system, including the inspiral, merger and ringdown (IMR), which became possible thanks to the breakthrough of numerical relativity~\cite{Pretorius05,Baker+06,Campanelli+06}, and the inclusion of the effects of the spins of the compact objects, that can lead to a precession of the orbital plane~\cite{ACST94,K95}. Besides accuracy, the computational efficiency of template generation is crucial.

@@

\jgb{To motivate our signal treatment we consider challenges of both the LISA response and GW precession applications in terms of a more general problem which covers both domains. In each case must determine the Fourier transform a signal $\tilde s(f)$ which results from processing some incident waveform $h(t)$ through some combination of modulations and delays.}

In the general problem, we apply a time-varying delay $d(t)$ to the waveform followed by a multiplicative modulation function $F(t)$,
\be
\label{eq:delay-mod-defs}
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,.
\ee
We then seek an \textit{efficient} way to compute the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee

As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form\cite{smoothAmpPhaseRefs}. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing us to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

In the case of precessing binaries, the delays will be absent and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from an \jgb{precessing orbital frame, in which waveform modes exhibit smooth amplitude and phase variation}, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector against the wave front, while the modulation will represent the time-variation in the detector orientation.

@@

Furthermore, a natural quantitative criterion for the applicability of this idea is given by the comparison of the radiation-reaction timescale of the signal with the modulation timescale. In other words, the change in frequency of the signal over a characeteristic time of the modulation should be small for the separation of timescales to work.

@@

Written in terms of the fractional laser frequency shifts between two spacecraft, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\begin{align}\label{eq:yslr}
	y_{slr} &\equiv \frac{\nu_{r} - \nu_{s}}{\nu} \nn\\
	&= \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hatk\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - L - \hatk\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hatk\cdot p_{r}) \right] \,.
\end{align}
We will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting point-ahead corrections). The additional delay $L$ (we use $c=1$) in the first term represents the light propagation time along the arm between spacecraft $s$ and $l$.

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. Sec.~\ref{subsec:modelLISA} provides more details on the response of LISA-type detectors, on the TDI observables, and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above.

@@

For the LISA response, the functions $F(t)$ and $d(t)$ vary on a timescale of one year, with a frequency $f_{0} = 1/\mathrm{yr} \simeq 3.169\times10^{-8} \mathrm{Hz}$ (we will also use $\Omega_{0} = 2\pi f_{0}$). When neglecting small non-periodic orbital perturbations (such as those caused by the influence of the Earth and of the other planets on the orbits), both $F(t)$ and $d(t)$ are periodic. It will also be useful to separate the delays in two types of terms: the first, $d_{0} = -\hatk\cdot p_{0}$, relates the waveform at the SSB to the waveform at the center of the constellation, whereas the second, $d_{c}$, represents the various delays between the spacecraft of the constellation. \jgb{Concretely, we assume the basic reference design parameters of the LISA mission\cite{LISA17} recently selected by the European Space Agency(ESA), with orbit radius $R=1\,\mathrm{au}$ and armlength $L=2.5\times10^{6}\mathrm{km}$, and with $d_{0} \sim R/c \simeq 500\mathrm{s}$ and $d_{c} \sim L/c \simeq 8\mathrm{s}$.}

@@

For spinning compact objects, \jgb{angular momentum interactions typically} lead to the precession of the orbit~\cite{Apostolatos+94, Kidder95}, which can have large effects on the waveform. In particular, it breaks the planar symmetry of the gravitational wave emission, causing modulations that are especially important for systems that are observed edge-on.

@@

If $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention, the modes in the inertial frame $h_{\ell m}^{\rm I}$ are then related to the modes in the precessing frame $h_{\ell m}^{\rm P}$ by~\cite{Goldberg+67}
\be\label{eq:wignerrotintro}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell *}_{mm'} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,,
\ee
where the $\calD^{\ell}_{mm'}$ are Wigner matrices~\cite{} \SM{[C]}. Notice that there is no mixing of the modes with different values of $\ell$.

If we make the assumption that the precessing-frame waveform $h^{\rm P}$ is approximated by a non-precessing model that provides us with a smooth Fourier-domain amplitude and phase, then the problem reduces to computing the Fourier transform of the signal
\be\label{eq:defmodulationprec}
	s(t) = F(t) h(t) \,,
\ee
where the modulation function $F(t)$ is given by a Wigner matrix and depends on time through the Euler angles $(\alpha, \beta, \gamma)(t)$. We will discuss more in Sec.~\ref{subsec:modelPrec} \SM{[R]} the limitations of the assumptions used to perform the decomposition above.

In Eq.~\eqref{eq:defmodulationprec} above, the modulation function $F$ has time variations on the precessional timescale, which evolves throughout the inspiral. The separation between the precessional timescale and the orbital timescale will be crucial for our analysis. In the limit of low frequencies, we will see in Sec.~\ref{} \SM{[R]} below that, although these timescales become more and more separated, the decrease in the chirping rate gives raise to a corrective contribution that does not vanish in this limit. In the other limit, as the system gets close to merger, the separation of timescales becomes weaker, and we will explore in Sec.~\ref{} \SM{[R]} the application of our formalism to the merger and ringdown phase.

@@

As a preliminary step, we recall here an approximation widely used for similar purposes in treating gravitational waves signals emitted by inspiraling binaries, the Stationary Phase Approximation (thereafter SPA, and also sometimes called the steepest-descent method). It applies in general to chirping signals, and we refer the reader to~\cite{FC93, CF94} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notation used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions\footnote{Note that the third condition is not always written explicitly, in particular not in Refs.~\cite{FC93, CF94}. It becomes important if we generalize $a$ to be an envelope function incorporating a modulation.} \SM{[say that the third is sometimes missing ?]} \SM{[add reference to Finn-Chernoff, Cutler-Flanagan where the third one is missing]}
\be\label{eq:conditionsSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(\tfSPA) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $\tfSPA$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(\tfSPA) \exp\left[2i\pi f \tfSPA-2i\varphi(\tfSPA) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (\tfSPA) (t-\tfSPA)^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{The expressions given here are valid for the dominant mode $h_{22}$ of the waveform. They can be generalized to other modes $h_{\ell m}$, $m\neq 0$ with phase $e^{-im\varphi}$ as follows: $\Psi$ acquires a factor $m/2$, $A$ a factor $\sqrt{2/m}$, and $v$ (see~\eqref{eq:SPAN}) has to be replaced by $(2/m)^{1/3} v$.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(\tfSPA) \sqrt{\frac{\pi}{\dot{\omega}(\tfSPA)}} \,, \label{eq:ASPA} \\
	\Psi_{\rm SPA}(f) &= 2\varphi(\tfSPA) - 2\pi f \tfSPA + \frac{\pi}{4} \,. \label{eq:PsiSPA}
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

@@

In this Newtonian, low-frequency limit one can check that each of the combinations \eqref{eq:conditionsSPA} indeed vanish at $\calO{(v^5)}$. but as the system approaches merger, these condition are no-longer satisfied. Near and after the merger, the SPA treatment is not applicable.
%, as can be seen by the breakdown of the relation~\eqref{eq:deftfSPA} as $\omega$ asymptotes the quasi-normal mode frequency in the ringdown. Because of this limitation, we aim at developing a formalism that would not rely on this approximation.
Nonetheless, the SPA has been a useful workhorse in many gravitational analyses involving frequency domain transfer functions of the form~\eqref{eq:deftransfer}. The usual approach is simply to replace any time dependencies appearing the transfer function using \eqref{eq:deftfSPA}. \jgb{We will reference the SPA treatment, as a familiar touchstone, as we develop a more general formalism.}

@@

A direct computation of the generalized convolution~\eqref{eq:FDkernel} using~\eqref{eq:defG} will generally be computationally demanding, but we can exploit the separation of timescales in the problem to seek an accurate but efficient approximation.  \jgb{We will explore approximations which allow the integral in~\eqref{eq:FDkernel} to be computed, so that the signal may be expressed the form of~\eqref{eq:deftransfer}, with the modulation and delay reduced to a multiplicative transfer function $\mathcal T(f)$.}

\jgb{For example, under appropriate conditions with slowly varying modulations and delays, we can expect that} the Fourier transform $\tilde{G}(f,f')$ should have a compact support, limited to $f' \in [-f_{\rm max}, f_{\rm max}]$ with $f_{\rm max}$ a maximal frequency for the modulation, roughly the inverse of its characteristic timescale. This makes the convolution integral~\eqref{eq:FDkernel} localized in frequency, as the waveform is to be evaluated only at frequencies close to $f$. We can then approximate $\tilde{h}(f-f')$ by $\tilde{h}(f)$, with a simple Taylor expansion for a perturbative treatment of the difference between the two. Doing so, we will recover at leading order a locality in time: the response is approximately reduced to an evaluation of the modulation and delay at a representative signal-dependent time $\tf$, which will be equivalent to $\tfSPA$ for inspiral signals.

@@

In presence of delays, the above is also complicated by the additional delay phases in the signal Fourier transform. Depending on the frequency $f$, this can lead $G(f,t)$ to have faster variations than its nominal timescale ($1\mathrm{yr}$ for LISA). As we will see in Sec.~\ref{} \SM{[R]}, \jgb{to derive useful approximations we will not only need to compare modulation timescales, but also to constrain dimensionless factors of the form $2\pi f d$.}

\jgb{In the next subsections we will examine the approximations necessary to allow a local-in-frequency response treatment, and consider low-order corrections which can improve accuracy in cases where the approriate conditions are more marginally realized.}

@@

\jgb{If the effective width of the kernel function $G(f,f’)$ is not quite negligible compared to the scale of significant variations with $f$ it can be useful to extend our approach beyond the leading-order approximation. With the waveform represented in phase-and-amplitude form~\eqref{eq:A9}, the elements of~\eqref{eq:FDkernel} may be Taylor expanded in the variable $f'$: \SM{[change notation, do not use $N$ as sum boundary for $\Psi$]}
\begin{subequations}\label{eq:expandfprime}
\begin{align}
	\Psi(f-f') &= \Psi(f) + 2\pi f' \tf + \sum\limits_{p\geq 2}^{N} \frac{(-1)^{p}}{p!} {f'}^{p} \frac{\ud^{p} \Psi}{\ud f^{p}} \,, \label{eq:expandPsi}\\
	A(f-f') &= A(f)+A(f) \sum\limits_{q\geq 1} \frac{(-1)^{q}}{q!} {f'}^{q} \frac{1}{A}\frac{\ud^{q} A}{\ud f^{q}} \,, \label{eq:expandA}\\
	\tilde{G}(f-f', f') &=\tilde G(f,f')+ \sum\limits_{r\geq 1} \frac{(-1)^{r}}{r!} {f'}^{r} \frac{\partial^{r} }{\partial f^{r}}  \tilde{G}(f,f') \label{eq:expandG} \,,
\end{align}
\end{subequations}
using the definition of $t_{f}$ introduced in~\eqref{eq:deftf}.}

\jgb{The leading order transfer function~\eqref{eq:transferlocal} is obtained by leaving off all the terms in the sums.  In the following, we will consider the resulting transfer functions when keeping some of the next few terms in each of these expansions.} 
Notice that we expand $\tilde{G}(f-f',f')$ in $f'$ only in its first argument, and that we can commute the $f$-derivatives of $G$ with the Fourier transform operation.

>>>>

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation, extending them by including delays and higher-order corrections while being applicable to a broader class of signals including full inspiral-merger-ringdown waveforms. In the LISA case, we find that higher-order corrections remain small except at the two extremes of the detector's frequency band, where we show that including them reduces the errors. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, higher-order corrections can become quantitatively large to the point of challenging our perturbative formalism, while affecting mainly the amplitude and giving rise to limited unfaithfulness. We further present an alternative direct convolution approach to accurately represent these post-merger features. \SM{[update to include SOBHs]}

@@

With the unprecedented recent gravitational-wave detections of coalescencing binary black holes and binary neutron stars, announced by the LIGO-Virgo collaboration~\cite{LIGO-theevent-2016, LIGO-christmasevent-2016, LIGO-O1BBH-2016, LIGO-BNSevent-2017}, gravitational-wave astronomy has entered its observational era. As LIGO prepares for even more sensitive observation runs, and with the recent expansion of the ground-based detectors network with Virgo~\cite{Virgo} (and eventually also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), observations of such compact object coalescences are expected at ever-increasing rate.

Moreover, the European Space Agency has recently selected the Laser Interferometer Space Antenna (LISA)\cite{LISA17} to realize the ``Gravitational Universe'' science theme\cite{elisa13} as the 3rd large space mission of its Cosmic Vision program, with a tentative launch around 2034. LISA will be able to detect and characterize, among several important gravitational wave source targets, comparable-mass binary black hole coalescences from cosmological distances over a wide range of masses. These will range from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

Data analysis for gravitational-wave observations of compact binary coalescences require accurate models (or templates) for the signals, both for ensuring efficient detections of signals that may be buried in instrumental noise, and to extract the physical parameters of the source in a subsequent analysis. Bayesian analysis for parameter estimation of gravitational-wave signals, as was performed for the LIGO detections~\cite{LIGO-theeventPE-2016,LIGO-O1BBH-2016}, may require millions of evaluations of the likelihood function to sample the posterior probability distribution. The greater sensitivity of LISA and other future instruments will require further increases in the accuracy and computational efficiency of signal templates.

@@

In our presentation, we will consider a formal signal processing problem encompassing the challenges posed by both the LISA response and the precession modulations. Given a signal $h(t)$, we apply a time-varying delay $d(t)$ to the waveform followed by a multiplicative modulation function $F(t)$,
\be
\label{eq:delay-mod-defs}
	h_{d}(t) = h(t+d(t)) \,, \quad s(t) = F(t)h_{d}(t) \,.
\ee
We then seek an efficient way to compute the Fourier transform $\tilde{s}(f)$, expressed by means of a Fourier-domain transfer function $\calT$ such that
\be\label{eq:deftransfer}
	\tilde{s}(f) \equiv \calT(f) \tilde{h}(f) \,. 
\ee

As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing us to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

In the case of precessing binaries, the delays will be absent and the modulation functions will be the time-dependent Wigner coefficients applied to rotate the waveform from a precessing orbital frame, in which waveform modes exhibit smooth amplitude and phase variation, to an inertial frame where the observations take place. In the case of a LISA-type detector, the signal $h(t)$ will simply be the waveform in a fixed heliocentric frame, and the delays will come from the motion of each detector against the wave front, while the modulation will represent the time-variation in the detector orientation.

@@

Furthermore, a natural quantitative criterion for the applicability of this idea is given by the comparison of the radiation-reaction timescale of the signal with the modulation timescale. In other words, the change in frequency of the signal over a characteristic time of the modulation should be large for the separation of timescales to work. However, as we will see below, for this problem the dimensional analysis falls short of the full picture: the separation of timescales can be affected by frequency-dependent dimensionless factors in presence of delays.

@@

Written in terms of the fractional laser frequency shifts between two spacecraft, the elementary response of the detector reads~\cite{EW75, RCP04, Vallisneri04}
\begin{align}\label{eq:yslr}
	y_{slr} &\equiv \frac{\nu_{r} - \nu_{s}}{\nu} \nn\\
	&= \frac{1}{2} \frac{n_{l}^{i}n_{l}^{j}}{1 - \hatk\cdot n_{l}} \left[ h_{ij}^{\rm TT}(t - L - \hatk\cdot p_{s}) - h_{ij}^{\rm TT}(t - \hatk\cdot p_{r}) \right] \,.
\end{align}
We will use a rigid instantaneous model, approximating the geometry of the constellation by a moving equilateral triangle (neglecting the flexing of the arms induced by corrections in the orbits), and evaluating all geometric factors in~\eqref{eq:yslr} at a single time $t$ (neglecting point-ahead corrections). The additional delay $L$ (we use $c=1$) in the first term represents the light propagation time along the arm between spacecraft $s$ and $r$.

The first- and second-generation TDI observables are then built as combinations of these basic building blocks, evaluated at delayed times. Since in our rigid approximation these delays will take a simple form in Fourier domain, we will focus in this paper on the $y_{slr}$ observables. Sec.~\ref{subsec:modelLISA} provides more details on the response of LISA-type detectors and on the approximations that enter the derivation of the basic response~\eqref{eq:yslr} above.

@@

For the LISA response, the functions $F(t)$ and $d(t)$ vary on a timescale of one year, with a frequency $f_{0} = 1/\mathrm{yr} \simeq 3.169\times10^{-8} \mathrm{Hz}$ (we will also use $\Omega_{0} = 2\pi f_{0}$). When neglecting small non-periodic orbital perturbations (such as those caused by the influence of the Earth and of the other planets on the orbits), both $F(t)$ and $d(t)$ are periodic. It will also be useful to separate the delays in two types of terms: the first, $d_{0} = -\hatk\cdot p_{0}$, relates the waveform at the SSB to the waveform at the center of the constellation, whereas the second, $d_{L}$, represents the various delays between the spacecraft of the constellation. Concretely, we assume the basic reference design parameters of the LISA mission\cite{LISA17} recently selected by the European Space Agency(ESA), with orbit radius $R=1\,\mathrm{au}$ and armlength $L=2.5\times10^{6}\mathrm{km}$, and with $d_{0} \sim R/c \simeq 500\mathrm{s}$ and $d_{L} \sim L/c \simeq 8\mathrm{s}$.

@@

For spinning compact objects, angular momentum interactions typically lead to the precession of the orbit~\cite{Apostolatos+94, Kidder95}, which can have large effects on the waveform. In particular, it breaks the planar symmetry of the gravitational wave emission, causing modulations that are especially important for systems that are observed edge-on.

@@

If $(\alpha, \beta, \gamma)$ are the Euler angles relating the precessing frame to the inertial frame in the $(z,y,z)$ convention, the modes in the inertial frame $h_{\ell m}^{\rm I}$ are then related to the modes in the precessing frame $h_{\ell m}^{\rm P}$ by~\cite{Goldberg+67}
\be\label{eq:wignerrotintro}
	h_{\ell m}^{\rm I} = \sum\limits_{m=-\ell}^{\ell} \calD^{\ell *}_{mm'} (\alpha,\beta,\gamma) h_{\ell m'}^{\rm P} \,,
\ee
where the $\calD^{\ell}_{mm'}$ are Wigner matrices (see App.~\ref{app:wigner}).

If we make the assumption that the precessing-frame waveform $h^{\rm P}$ is approximated by a non-precessing model that provides us with a smooth Fourier-domain amplitude and phase, then the problem reduces to computing the Fourier transform of the signal
\be\label{eq:defmodulationprec}
	s(t) = F(t) h(t) \,,
\ee
where the modulation function $F(t)$ is given by a Wigner matrix and depends on time through the Euler angles $(\alpha, \beta, \gamma)(t)$.

In Eq.~\eqref{eq:defmodulationprec} above, the modulation function $F$ has time variations on the precessional timescale, which evolves throughout the inspiral. In the limit of low frequencies, we will see in Sec.~\ref{sec:precession} below that, although the precession and orbital timescales become more and more separated, the decrease in the chirping rate gives raise to a corrective contribution that does not vanish in this limit. We will also explore  the application of our formalism to a precessing-frame decomposition of the waveform extended through the merger and ringdown phase.

@@

As a preliminary step, we recall here an approximation widely used for similar purposes in treating gravitational waves signals emitted by inspiraling binaries, the Stationary Phase Approximation (thereafter SPA, and also sometimes called the steepest-descent method). It applies in general to chirping signals, and we refer the reader to~\cite{FC93, CF94} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notation used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions\footnote{Note that the third condition is not always written explicitly, in particular not in Refs.~\cite{FC93, CF94}. It becomes important if we generalize $a$ to be an envelope function incorporating a modulation.}
\be\label{eq:conditionsSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(\tfSPA) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $\tfSPA$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(\tfSPA) \exp\left[2i\pi f \tfSPA-2i\varphi(\tfSPA) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (\tfSPA) (t-\tfSPA)^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{The expressions given here are valid for the dominant mode $h_{22}$ of the waveform. They can be generalized to other modes $h_{\ell m}$, $m\neq 0$ with phase $e^{-im\varphi}$ as follows: $\Psi$ acquires a factor $m/2$, $A$ a factor $\sqrt{2/m}$, and $v$ (see~\eqref{eq:SPAN}) has to be replaced by $(2/m)^{1/3} v$.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(\tfSPA) \sqrt{\frac{\pi}{\dot{\omega}(\tfSPA)}} \,, \label{eq:ASPA} \\
	\Psi_{\rm SPA}(f) &= 2\varphi(\tfSPA) - 2\pi f \tfSPA + \frac{\pi}{4} \,. \label{eq:PsiSPA}
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

@@

In this Newtonian, low-frequency limit one can check that each of the combinations \eqref{eq:conditionsSPA} indeed vanish at $\calO{(v^5)}$. but as the system approaches merger, these condition are no-longer satisfied. Near and after the merger, the SPA treatment is not applicable.
%, as can be seen by the breakdown of the relation~\eqref{eq:deftfSPA} as $\omega$ asymptotes the quasi-normal mode frequency in the ringdown. Because of this limitation, we aim at developing a formalism that would not rely on this approximation.
Nonetheless, the SPA has been a useful workhorse in many gravitational analyses involving frequency domain transfer functions of the form~\eqref{eq:deftransfer}. The usual approach is simply to replace any time dependencies appearing the transfer function using \eqref{eq:deftfSPA}. We will reference the SPA treatment, as a familiar touchstone, as we develop a more general formalism.

@@

A direct computation of the generalized convolution~\eqref{eq:FDkernel} using~\eqref{eq:defG} will generally be computationally demanding, but we can exploit the separation of timescales in the problem to seek an accurate but efficient approximation and to compute the transfer function $\calT(f)$ as in~\eqref{eq:deftransfer}.

For example, under appropriate conditions with slowly varying modulations and delays, we can expect that the Fourier transform $\tilde{G}(f,f')$ should have a compact support, limited to $f' \in [-f_{\rm max}, f_{\rm max}]$ with $f_{\rm max}$ a maximal frequency for the modulation, roughly the inverse of its characteristic timescale. This makes the convolution integral~\eqref{eq:FDkernel} localized in frequency, as the waveform is to be evaluated only at frequencies close to $f$. We can then approximate $\tilde{h}(f-f')$ by $\tilde{h}(f)$, with a simple Taylor expansion of their difference. Doing so, we will recover at leading order a locality in time: the response is approximately reduced to an evaluation of the modulation and delay at a representative signal-dependent time $\tf$, which will be equivalent to $\tfSPA$ for inspiral signals.

@@

In presence of delays, the above is also complicated by the additional delay phases in the signal Fourier transform. Depending on the frequency $f$, this can lead $G(f,t)$ to have faster variations than its nominal timescale ($1\mathrm{yr}$ for LISA). Thus, to assess the validity of our approximations we cannot limit ourselves to comparing the dimensionful timescales at play in the problem. As we will see in Sec.~\ref{subsec:lisafom}, we must also include in the analysis dimensionless factors of the form $2\pi f d$.

@@

If the width of the kernel function $G(f-f',f')$ is not quite negligible compared to the scale of significant variations of $\tilde{h}(f)$ with $f$, it can be useful to extend our approach beyond the leading-order approximation. With the waveform represented in the amplitude and phase form~\eqref{eq:defAPsi}, the elements of~\eqref{eq:FDkernel} may be formally Taylor-expanded in the variable $f'$:
\begin{subequations}\label{eq:expandfprime}
\begin{align}
	\Psi(f-f') &= \Psi(f) + 2\pi f' \tf + \sum\limits_{p\geq 2} \frac{(-1)^{p}}{p!} {f'}^{p} \frac{\ud^{p} \Psi}{\ud f^{p}} \,, \label{eq:expandPsi}\\
	A(f-f') &= A(f)+A(f) \sum\limits_{q\geq 1} \frac{(-1)^{q}}{q!} {f'}^{q} \frac{1}{A}\frac{\ud^{q} A}{\ud f^{q}} \,, \label{eq:expandA}\\
	\tilde{G}(f-f', f') &=\tilde G(f,f')+ \sum\limits_{r\geq 1} \frac{(-1)^{r}}{r!} {f'}^{r} \frac{\partial^{r} }{\partial f^{r}}  \tilde{G}(f,f') \label{eq:expandG} \,,
\end{align}
\end{subequations}
using the definition of $t_{f}$ introduced in~\eqref{eq:deftf}.

The leading order transfer function~\eqref{eq:transferlocal} is obtained by leaving off all the terms in the sums.  In the following, we will consider the resulting transfer functions when keeping some of the next few terms in each of these expansions. Notice that we expand $\tilde{G}(f-f',f')$ in $f'$ only in its first argument, and that we can commute the $f$-derivatives of $G$ with the Fourier transform operation.
